---
title: "Data import"
output: github_document
---

```{r}
library(tidyverse)
library(readxl)
library(haven)
```

Let's import a dataset. 

```{r}
  litters_df = 
    read_csv("data/FAS_litters.csv")
```

Look at your data !! 

```{r}
names(litters_df)
```

Update the names in 'litters_df'.

```{r}
litters_df=
    janitor::clean_names(litters_df)
```

Look at the data !!! 

```{r}
litters_df
```


Sometimes skimming data is neat ? 

```{r}
  skimr::skim(litters_df)
```


## Fix the missingness 

```{r}
  litters_df=
    read_csv("data/FAS_litters.csv", na = c("NA",".",""))
```


##Try uploading the pups data set and skip to three becasue that's where the data is 

```{r}
  pups_df= 
    read_csv(
      "data/FAS_pups.csv", 
      skip = 3,
      na = c("NA",".","")
      )

pups_df= 
  janitor::clean_names(pups_df)
```


## okay what about excel 

  CSVs are really great but sometimes you get an excel file. 
  
```{r}
mlb_df=
  read_excel("data/mlb11.xlsx")
```

Import LotR wordcounts. 

```{r}
  fotr_df= 
    read_excel("data/LotR_Words.xlsx", range = "B3:D6")
```


##SAS???


Import the PULSE data set 

```{r}
pulse_df=
  read_sas("data/public_pulse_data.sas7bdat")

pulse_df=
  janitor::clean_names(pulse_df)
```




## Why do I hate read.csv so much 

```{r, eval=FALSE}
litters_df_base=
  read.csv("data/FAS_litters.csv")
```

